A Spectral Analysis of the Domain Decomposed Adjoint Neumann-Ulam Method

### 2012-11-21 doc/mandc2013/notes/spectral_theory.tex
uadded convergence section

For reactor physics Monte Carlo simulations, domain decomposition has
been identified as a key principle in moving forward in high
performance computing
\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progress and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain. We wish
to characterize this communication not only because communication is
in general expensive, but also because nearest-neighbor communication
sequences have poor algorithmic strong scaling \cite{gropp_cfd_2001}.

@@

%%---------------------------------------------------------------------------%%
\section{The Adjoint Neumann-Ulam Method}
We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a linear operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
the forcing term. Choosing the adjoint method Monte Carlo method to
invert the linear operator, we begin by defining the linear system
adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
respectively and $\ve{A}^T$ is the adjoint operator.  With this
statement we can then define the split equation:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H}$ is defined as the \textit{iteration matrix}.  For
convergence, we require that the spectral radius of $\ve{H}$ must
remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
therefore has the same spectral radius.  Using these definitions, we
can derive an estimator from the adjoint method that will also give
the solution vector, $\ve{x}$. As with the direct method, we can
acquire the adjoint solution by forming the Neumann series by writing
Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence, this time
forming the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the direct solution. The
adjoint estimator can be related to the solution by defining the
following inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
From this definition it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
 Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$ and therefore we
 require two constraints to close the system. We use
 Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as
 a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_i\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
the Dirac delta function $\delta_{i_k,i}$. If we apply
Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_i \rangle = x_i \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the direct source and
the adjoint solution using a delta function source, we recover the
direct solution we are looking for

@@

Like the direct method, we also desire a criteria for random walk
termination for problems where only an approximate solution is
necessary. For the adjoint method, we utilize a \textit{relative
  weight cutoff}:
\begin{equation}
  W_f = W_c b_{i_0}\:,
  \label{eq:relative_weight_cutoff}
\end{equation}
where $W_c$ is defined as in the direct method. The adjoint random
walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
contributions become increasingly small.

%%---------------------------------------------------------------------------%%
\section{Model Problem}
For our numerical experiments, we choose the one-speed,
two-dimensional neutron diffusion equation as the model problem
\cite{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogenous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
\cite{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.25}{\input{stencil.pdftex_t}}
  \end{center}
  \caption{\textbf{Nine-point Laplacian stencil.}}
  \label{fig:stencil}
\end{figure}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{S}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unkown fluxes. We
note here that the diffusion operator is symmetric.

@@

%%---------------------------------------------------------------------------%%
\section{Spectral Analysis}
The convergence of the Neumann series approximated by the Monte Carlo
solver is dependent on the Eigenvalues of the iteration matrix. In
addition, the Eigenvalues of the linear operator from which the
iteration matrix was derived allow for further analysis of convergence
properties. We will compute the Eigenvalues of these matrices by
assuming Eigenfunctions of the form \cite{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
Eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these Eigenfunctions should
give the Eigenvalues of the matrix as they lie on the unit circle in
the complex plane.

\subsection{Diffusion Operator Spectrum}
For the model problem, we first compute the Eigenvalues for the
diffusion operator $\ve{D}$ by applying the operator the
Eigenfunctions and noting that $x=ih$ and $y=jh$:
\begin{multline}
  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D}) =\\ -\frac{D}{6h^2}[4
    e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi
      \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2
      \pi \imath q h} \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} +
    e^{2 \pi \imath p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h}
    e^{2 \pi \imath q h} - 20] + \Sigma_a \:.
  \label{eq:deriv_diff_1}
\end{multline}
Using Euler's formula, we can collapse the exponentials to
trigonometric functions:
\begin{equation}
  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
  \label{eq:deriv_diff_2}
\end{equation}

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi
preconditioning is sufficient to reduce the spectral radius of the
operator below unity and therefore ensure convergence of the Neumann
series. The preconditioner in this case is then $\ve{M} =
diag(\ve{D})$ such that we are solving the following linear system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
operator with each row scaled by the diagonal component. As we have
defined a homogenous domain, this scaling factor is the same for all
rows in the operator. This scaling factor, $\alpha$, is then defined
as the $\phi_{i,j}$ coefficient from Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = [\frac{10 D}{3 h^2} + \Sigma_a]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we then have the following spectrum of
preconditioned eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

A good measure of how well-conditioned a linear system is can be
obtained by computing the condition number for the linear operator
defined as:
\begin{equation}
  \kappa(\ve{A}) = ||\ve{A}||\ ||\ve{A}^{-1}||\:,
  \label{eq:condition_number}
\end{equation}
which can be related to the Eigenvalues of the operator as:
\begin{equation}
  \kappa(\ve{A}) =
  \frac{\max_{p,q}\lambda_{p,q}(\ve{A})}{\min_{p,q}\lambda_{p,q}(\ve{A})}\:.
  \label{eq:condition_number_2}
\end{equation}
To compute the condition number for the preconditioned diffusion
system, we must therefore compute the minimum and maximum Eigenvalues
of the diffusion operator. Per Leveque's text
\cite{leveque_finite_2007}, we expect to find these minimum and
maximum eigenvalues when $p=q$. To find these values,
Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
values of $\Sigma_a$.

We see a minimum eigenvalue at $p=0$ and a maximum at $p=10$, letting
us compute the condition number for the Jacobi preconditioned
diffusion operator:
\begin{equation}
  \kappa(\ve{M}^{-1} \ve{D}) =
  \frac{\frac{D}{6h^2}\Sigma_a}{\frac{D}{3h^2}[8 \cos(\pi 10 h) + 2
      \cos^2(\pi 10 h) - 10] + \Sigma_a}\:.
  \label{eq:diffusion_condition}
\end{equation}
Visually we can also see how increasing the absorption in the problem
creates a more well conditioned system as less diffusion is
expected. If we have a problem with only scattering, $\Sigma_a = 0$
and we observe an infinite condition number as $\lambda_{min}
\rightarrow 0$. We note also that the diffusion operator contains
purely positive Eigenvalues and can therefore be classified as
symmetric positive definite (SPD).

\subsection{Iteration Matrix Spectrum}

\subsection{Neumann Series Convergence}

\subsection{The Wigner Rational Approximation}

>>>>

For reactor physics Monte Carlo simulations, domain decomposition has
been identified as a key principle in moving forward in high
performance computing
\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progresses and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain. We wish
to characterize this communication not only because communication is
in general expensive, but also because nearest-neighbor communication
sequences have poor algorithmic strong scaling \cite{gropp_cfd_2001}.
We observe this because nearest neighbor communication patterns scale
with the volume and surface sizes of the local domains. The larger the
volume, the more compute time is spent within the local volume and
therefore at some point you are limited by the latency of the system,
or the time it takes to receive the data to operate on. However, at a
certain point, when the surface of those volumes becomes large enough,
the amount of data that must be communicated grows large enough that
it may overcome the bandwidth limitations of the system, resulting in
a decrease in performance.

@@

%%---------------------------------------------------------------------------%%
\section{The Adjoint Neumann-Ulam Method}
We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a linear operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
the forcing term. Choosing the adjoint method Monte Carlo method to
invert the linear operator, we begin by defining the linear system
adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
respectively and $\ve{A}^T$ is the adjoint operator.  With this
statement we can then define the split equation:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H}$ is defined as the \textit{iteration matrix}.  Using
these definitions, we can derive an estimator from the adjoint method
that will also give the solution vector, $\ve{x}$. We can acquire the
adjoint solution by forming the Neumann series by writing
Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
For this series to converge, we require that the spectral radius of
$\ve{H}$ must remain less than 1 as $\ve{H}^T$ contains the same
eigenvalues and therefore has the same spectral radius. We expand this
summation to again yield a series of transitions that can be
approximated by a Monte Carlo random walk sequence, this time forming
the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the direct solution. The
adjoint estimator can be related to the solution by defining the
following inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
From this definition it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$ and therefore we
require two constraints to close the system. We use
Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as
a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_i\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
the Dirac delta function $\delta_{i_k,i}$. If we apply
Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_i \rangle = x_i \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the direct source and
the adjoint solution using a delta function source, we recover the
direct solution we are looking for

@@

We also desire a criteria for random walk termination for problems
where only an approximate solution is necessary. For the adjoint
method, we utilize a \textit{relative weight cutoff}:
\begin{equation}
  W_f = W_c b_{i_0}\:,
  \label{eq:relative_weight_cutoff}
\end{equation}
where $W_c$ is defined as in the direct method. The adjoint random
walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
contributions become increasingly small.

%%---------------------------------------------------------------------------%%
\section{Model Problem}
For our numerical experiments, we choose the one-speed,
two-dimensional neutron diffusion equation as a model problem
\cite{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, and $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogenous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
\cite{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.25}{\input{stencil.pdftex_t}}
  \end{center}
  \caption{\textbf{Nine-point Laplacian stencil.}}
  \label{fig:stencil}
\end{figure}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unkown fluxes. We
note here that the diffusion operator is symmetric.

@@

%%---------------------------------------------------------------------------%%
\section{Spectral Analysis}
The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the Eigenvalues of the iteration matrix. In
addition, the Eigenvalues of the linear operator from which the
iteration matrix was derived allow for further analysis of convergence
properties. We will compute the Eigenvalues of these matrices by
assuming Eigenfunctions of the form \cite{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
Eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these Eigenfunctions should
give the Eigenvalues of the matrix as they lie on the unit circle in
the complex plane.

\subsection{Diffusion Operator Spectrum}
For the model problem, we first compute the Eigenvalues for the
diffusion operator $\ve{D}$ by applying the operator to the
Eigenfunctions and noting that $x=ih$ and $y=jh$:
\begin{multline}
  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D}) =\\ -\frac{D}{6h^2}[4
    e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi
      \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2
      \pi \imath q h} \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} +
    e^{2 \pi \imath p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h}
    e^{2 \pi \imath q h} - 20] + \Sigma_a \:.
  \label{eq:deriv_diff_1}
\end{multline}
Using Euler's formula, we can collapse the exponentials to
trigonometric functions:
\begin{equation}
  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
  \label{eq:deriv_diff_2}
\end{equation}

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi
preconditioning is sufficient to reduce the spectral radius of the
iteration matrix below unity and therefore ensure convergence of the
Neumann series. The preconditioner in this case is then $\ve{M} =
diag(\ve{D})$ such that we are solving the following linear system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
operator with each row scaled by the diagonal component. As we have
defined a homogenous domain, the scaling factor is the same for all
rows in the operator. This scaling factor, $\alpha$, is then defined
as the $\phi_{i,j}$ coefficient from Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = [\frac{10 D}{3 h^2} + \Sigma_a]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we then have the following spectrum of
preconditioned eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

A good measure of how well-conditioned is a linear system can be
obtained by computing the condition number for the linear operator
defined as:
\begin{equation}
  \kappa(\ve{A}) = ||\ve{A}||\ ||\ve{A}^{-1}||\:,
  \label{eq:condition_number}
\end{equation}
which can be related to the Eigenvalues of the operator as:
\begin{equation}
  \kappa(\ve{A}) =
  \frac{\max_{p,q}\lambda_{p,q}(\ve{A})}{\min_{p,q}\lambda_{p,q}(\ve{A})}\:.
  \label{eq:condition_number_2}
\end{equation}
To compute the condition number for the preconditioned diffusion
system, we must therefore compute the minimum and maximum Eigenvalues
of the diffusion operator. Per Leveque's text
\cite{leveque_finite_2007}, we expect to find these minimum and
maximum eigenvalues when $p=q$. To find these values,
Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
values of $\Sigma_a$.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{diffusion_spectrum.png}
  \end{center}
  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
  \label{fig:diffusion_spectrum}
\end{figure}
We see a minimum eigenvalue at $p=0$ and a maximum at $p=10$, letting
us compute the condition number for the Jacobi preconditioned
diffusion operator:
\begin{equation}
  \kappa(\ve{M}^{-1} \ve{D}) =
  \frac{\frac{D}{6h^2}\Sigma_a}{\frac{D}{3h^2}[8 \cos(10 \pi h) + 2
      \cos^2(10 \pi h) - 10] + \Sigma_a}\:.
  \label{eq:diffusion_condition}
\end{equation}
Visually we can also see how increasing the absorption in the problem
creates a more well conditioned system as less diffusion is
expected. If we have a problem with only scattering, $\Sigma_a = 0$
and we observe an infinite condition number as $\lambda_{min}
\rightarrow 0$. We note also that the diffusion operator contains
purely positive Eigenvalues and can therefore be classified as
symmetric positive definite (SPD).

\subsection{Iteration Matrix Spectrum}
The spectral radius of the iteration matrix is obtained by seeking its
largest Eigenvalue. As with the diffusion operator, we can use the
same analysis techniques to find the Eigenvalues for the iteration
matrix. We use a few simplifications by noting that if the Jacobi
preconditioned spectral radius is $\ve{H} = \ve{I} -
\ve{M}^{-1}\ve{D}$, the we expect all terms on the diagonal of the
iteration matrix to be zero such that we have the following stencil:
\begin{equation}
  \ve{H}\boldsymbol{\phi} = -\frac{\alpha D}{6h^2}[4 \phi_{i-1,j} + 4
    \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1}]\:.
  \label{eq:iteration_stencil}
\end{equation}
Inserting the Eigenfunctions defined by Eq~\ref{eq:eigenfunction_form}
we get:
\begin{multline}
  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}[4 e^{-2 \pi \imath p
      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
      \imath q h}]\:,
  \label{eq:iteration_deriv}
\end{multline}
which simplifies to:
\begin{equation}
  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
  \label{eq:iteration_spectrum}
\end{equation}
giving the Eigenvalue spectrum for the Jacobi preconditioned iteration
matrix. As with the linear operator, we expect the maximum Eigenvalue
to exist when $p=q$. In addition, we note that we have the same cosine
form of the spectrum as for the linear operator expect that the
spectra will be shifted by 10 in the $-p$ direction. We therefore
expect the maximum Eigenvalues to be found when $p=q=0$, giving the
following for the spectral radius of the Jacobi precondition iteration
matrix:
\begin{equation}
  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
  \label{eq:iteration_radius}
\end{equation}

\subsection{Neumann Series Convergence}
The Neumann-Ulam method for matrix inversion is effectively an
approximation to a stationary method (the Jacobi method in the case
of the preconditioning above). Stationary methods for linear systems arise from splitting the
operator in Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A} = \ve{M} - \ve{N}\:,
  \label{eq:split_linear_operator}
\end{equation}
where the choice of $\ve{M}$ and $\ve{N}$ will be dictated by the
particular method chosen. Using this split definition of the operator
we can then write:
\begin{equation}
  \ve{M}\ve{x} - \ve{N}\ve{x} = \ve{b}\:.
  \label{eq:linear_split_equation1}
\end{equation}
By rearranging, we can generate a form more useful for analysis:
\begin{equation}
  \ve{x} = \ve{H}\ve{x} + \ve{c}\:,
  \label{eq:linear_split_equation2}
\end{equation}
where $\ve{H}=\ve{M}^{-1}\ve{N}$ is defined as the \textit{iteration
  matrix} and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
both the left and right hand sides, an iterative method can then be
formed:
\begin{equation}
  \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbb{Z}^+$ defined as the \textit{iteration
  index}. Given this, we can then generate a few statements regarding
the convergence of such stationary methods. Defining $\ve{e}^k =
\ve{u}^k - \ve{u}$ as the solution error at the $k^{th}$ iterate, we
can subtract Eq~(\ref{eq:linear_split_equation2}) from
Eq~(\ref{eq:linear_iterative_method}) to arrive at an error form of
the linear problem:
\begin{equation}
  \ve{e}^{k+1} = \ve{H}\ve{e}^k\:. 
  \label{eq:linear_iterative_error}
\end{equation}
Our error after $k$ iterations is then:
\begin{equation}
  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
  \label{eq:linear_k_iter_error}
\end{equation}
In other words, successive application of the iteration matrix is the
mechanism driving down the error in a stationary method. We can then
place restrictions on the iteration matrix by assuming $\ve{H}$ is
diagonalizable \cite{saad_iterative_2003}, we then have:
\begin{equation}
  \ve{e}^{k} =
  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\ve{e}^0\:,
  \label{eq:linear_k_iter_error_diag}
\end{equation}
where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\ve{H}$ on
its diagonal and the columns of $\ve{R}$ contain the Eigenvectors of
$\ve{H}$. Computing the 2-norm of the above form then gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\ve{e}^0||_2\:,
  \label{eq:linear_k_iter_norm1}
\end{equation}
which gives:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
  ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm2}
\end{equation}
For iteration matrices where the Eigenvectors are orthogonal,
$\kappa(\ve{R})=1$ and the error bound reduces to:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}
We can now restrict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
for a stationary method to converge such that $k$ applications of the
iteration matrix will not cause the error to grow in
Eq~(\ref{eq:linear_k_iter_norm3}).
